\bf An HMM-based part-of-speech tagger for Turkish

### 2015-12-09 paper/paper.tex
Almost complete

\section{Results}
The POS tagger I've built as proposed can tag words with a
reasonable accuracy which we talk about in \ref{subsec:accuracy}.

@@

\paragraph{\texttt{create-model}}
The executable \texttt{create-model} is implemented in two files:
\texttt{Parse.hs} and \texttt{Main.hs}. \texttt{Parse.hs} handles the parsing of
the corpus.

@@

\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
freqMap :: Ord a $\Rightarrow$ [a] $\rightarrow$ M.Map a Int
freqMap unigrams = populate M.empty unigrams
  where populate :: Ord a M.Map a $Z$ $\rightarrow$ [a] $\rightarrow$ M.Map a Int
        populate m [] = m
        populate m (x:xs) = let freq = (M.findWithDefault 0 x m) :: ℤ
                            in populate (M.insert x (freq + 1) m) xs
\end{lstlisting}
Once the required lists are generated using these they are saved to a file in
the directory \texttt{model}.

\paragraph{\texttt{tag}}

The executable \texttt{run-tag} is where the actual tagging happens. The
frequencies we have pre-computed in \texttt{create-model} are passed in to the
probability function which we give below.
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
probability :: (Ord a, Ord b) $\Rightarrow$ (a, b) $\rightarrow$ Map a Int $\rightarrow$ Map (a, b) ℤ $\rightarrow \mathscr{L}$
probability (x, y) c$_1$ c$_2$ = if xCount == 0 || yCount == 0
                                   then logFloat $\epsilon$
                                   else (logFloat yCount) / (logFloat xCount)
  where yCount = fromIntegral (findWithDefault 0 (x, y) c₂) ∷ Double
        xCount = fromIntegral (findWithDefault 0 x c₁) ∷ Double
\end{lstlisting}

\texttt{probability} when combined with \texttt{freqMap} yield an efficient
basis for implementing Bayes' law. When we are creating the functions
required by \texttt{HMM} we are storing the map returned by a call to
\texttt{freqMap} in the lexical closure of \texttt{probability} by making use
Haskell's currying so that we can eloquently define the two crucial functions
as follows:
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
transFn s$_1$ s$_2$ = probability (s$_1$, s$_2$) tagFreqs tagBigramFreqs
outFn s e = probability (e, s) wordFreqs taggedWordFreqs
\end{lstlisting}

@@

I have tested the POS tagger with 6075 words from a small portion of the corpus
that I reserved for testing (i.e., did not use for tranining the HMM). The POS
tagger tagged 4514 of these words correctly (74.3 \%). Here is a brief list of
known problems:
\begin{description}
  \item[Duplications:] The Turkish language contains lots of duplications.
  Syntactically speaking, a duplication behaves like an word itself. In the
  treebank, reduplication are denoted as as single word composed of joining the
  two halves of the duplication with an underscore (half$_1$\_$half_2$). Since
  there is not an easy way to detect whether two sequential words are
  reduplications or not, I am giving duplications in an incorrect format to the
  HMM. I am predicting (based on my informal tests with the model) that
  correction of this problem would increase the accuracy by about $\%5$.
  \item[Relaxed word order:] Turkish is generally considered a free constituent
  order language \citep{erguvanli, dincer2008suffix}, although there does not seem to be an
  agreed-upon view on this. Yet, it would be safe to say that the word order of
  Turkish is more relaxed compared to English. The HMM-based approach to POS
  tagging presupposes the ordered-ness of the language. Although there is a
  canonical word order in Turkish, it tends not to be rigidly followed all the
  time---the morphosyntax of Turkish allows variation. It might be possible to
  increase the accuracy of the HMM by, for example, isolating bigrams in
  dependent clauses from those in their parents might work since Turkish, due to
  its relaxed order, permits lots of dependent clauses.
\end{description}

@@

The hidden markov model I've built for this project \emph{ignores} morphology.

% Delete this todo
\todo[inline]{Write a concluding summary. Also indicate what future work may be warranted. This section should be at least half a page in length.}

>>>>

\section{Results}
The resulting program I have implemented is based on bigram hidden markov
models. To summarize it we are making use of three probabilities: (1) $P(t_i |
t_{i-1})$ the probability of being at a state (i.e., a tag) given the previous
state. This is computed\footnote{Although I use the word ``compute'', everything
  is already pre-computed and the referred computation is merely a lookup. This
  is discussed in \ref{par:tag}} by the function \texttt{transFn}. (2) $P(w_i | t_i)$ the
probability of seeing a word having a given tag, computed by the function
\texttt{outFn}. Finally, the probability of a given state being the start state
of a sentence, computed by the function \texttt{outProbs}.

@@

\subsubsection{\texttt{create-model}}
The executable \texttt{create-model} is implemented in two files:
\texttt{Parse.hs} and \texttt{Main.hs}. \texttt{Parse.hs} handles the parsing of
the corpus.

@@

\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
freqMap :: Ord a $\Rightarrow$ [a] $\rightarrow$ M.Map a Int
freqMap elems = populate empty elems
  where populate :: Ord a Map a Int $\rightarrow$ [a] $\rightarrow$ M.Map a Int
         populate m [] = m
         populate m (x:xs) = let freq = (M.findWithDefault 0 x m) :: Int
                            in populate (M.insert x (freq + 1) m) xs
\end{lstlisting}

This function is called several times with lists of values we want to
\emph{count} such as tag bigrams, a given word with a certain tag et cetera.
Pre-computing the counts is almost mandatory since frequencies (i.e., counts)
will need to be accessed frequently when computing the probabilities. Once the
necessary maps are generated, they are written to files in the directory
\texttt{model}.

\subsubsection{\texttt{tag}}
\label{par:tag}
The executable \texttt{run-tag} is where the actual tagging happens. The
frequencies we have pre-computed in \texttt{create-model} are passed in to the
probability function which we give below.
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
probability :: (Ord a, Ord b) $\Rightarrow$ (a, b) $\rightarrow$ Map a Int $\rightarrow$ Map (a, b) ℤ $\rightarrow \mathscr{L}$
probability (x, y) c$_1$ c$_2$ = if xCount == 0 || yCount == 0
                                   then logFloat $\epsilon$
                                   else (logFloat yCount) / (logFloat xCount)
  where yCount = fromIntegral (findWithDefault 0 (x, y) c₂) ∷ Double
        xCount = fromIntegral (findWithDefault 0 x c₁) ∷ Double
\end{lstlisting}

This is a straightforward implementation of probability relying on a given count
map. One thing to note is that it returns $\epsilon$ probability (i.e., the
smalles value a \texttt{logFloat}) if a given element is not found.

\texttt{probability} when combined with \texttt{freqMap} forms an
efficient method for implementing Bayes' law. When we are creating the functions
required by \texttt{HMM} we are storing the map returned by a call to
\texttt{freqMap} in the lexical closure of \texttt{probability} by making use
Haskell's currying so that we can eloquently define the two crucial functions as
follows:
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
transFn s$_1$ s$_2$ = probability (s$_1$, s$_2$) tagFreqs tagBigramFreqs
outFn s e = probability (e, s) wordFreqs taggedWordFreqs
\end{lstlisting}
The probability functions used to create \texttt{transFn} and \texttt{outFn} hold
the count maps \texttt{tagBigramFreqs} and \texttt{taggedWordFreqs} in their
lexical closure, respectively.

Finally we declare a new \texttt{HMM} as follows:

\newpage
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
      newHMM = HMM { states      = possibleTags ∷ [POS]
                     , events      = ws ∷ [String]
                     , initProbs   = initProbFn
                     , transMatrix = transFn
                     , outMatrix   = outFn
                     }
\end{lstlisting}

This is a simple record (a record is like C struct in Haskell) containing the
probability functions we defined.\footnote{For some reason, the author of the
  \texttt{Data.HMM} library named the fields (i.e., attribute), that are of
  function types, using names that end with \texttt{-Matrix} although they are
  \emph{not} matrices. The reader should not be confused by this.}

We have shown that it is possible---though not ideal---to build a POS tagger for
Turkish that is based entirely on a bigram hidden markov model. Let us first
look at the accuracy of this model and propose possible fixes that do not
fundamentally change the structure of the model.

@@

We have tested the POS tagger with 6075 words from a small portion of the corpus
that I reserved for testing (i.e., did not use for tranining the HMM). The POS
tagger tagged 4514 of these words correctly (74.3 \%). Here is a brief list of
known problems:

\begin{description}

  \item[Duplications:] The Turkish language contains many duplications.
  Syntactically speaking, a duplication behaves like a word itself. In the
  treebank, duplications are denoted with two halves joint the with an
  underscore. Since there is not an easy way to detect whether two sequential
  words are reduplications or not, we are giving duplications in an incorrect
  format to the HMM.  We predict that (based on informal tests with the model)
  that correction of this problem would significantly increase the accuracy.

  \item[Abundance of independent clauses:] It is common in the Turkish language
  to have multiple independent clauses in a given sentence. It is possible that
  this decreases the accuracy of the HMM: instead of a tag $t$ being followed by
  a tag $t'$, we get a lot of $t, <clause>, t'$ which obscures the word order we
  are interested in. Since we may always assign the \textsc{Punc} tag to a comma
  with full certainty, we may simply ignore commas as far as probability is
  concerned and tag them with \textsc{Punc}.

\end{description}

@@

\subsection{Further research}

A salient possibility for further research is the incorporation of word endings
into the model using regular expressions. It is trivial to express Turkish
suffixes, that change their form as not to upset the major vowel harmony in a
given word. We may say with probably more than $90\%$ probability that a given
word has a certain POS looking whether its ending satisfies regular expressions.
